---
title: "00.5_forecast_data_prep"
author: "JTK"
date: "2025-05-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Housekeeping

### Packages

```{r, eval=FALSE}

# Data mgmt
require(tidyverse)

## Parallel computing
require(future)
require(furrr)

```

# Pull-in and manipulate forecasts

### Load from file
```{r}

#### Set forecast file location and list all the files located there

loc <- here("output-data/nwm_operational/medium_term/")

mt_files <- list.files(loc)

mt_downloads <- list()

#### Import all the downloaded files

for(i in 1:length(mt_files)){
  
  file_name <- here(loc, mt_files[i])
  
  mt <- read_csv(file_name)
  
  mt_downloads[[i]] <- mt 
  
  
}

#### Trim down to Lake Champlain gages

mt_downloads_lc <- map(mt_downloads, ~(.x %>%
                                         filter(as.character(comid) %in% 
                                                  lc_gages_metadata_clean$comid)),
                       .progress = TRUE)

#### Make tibble

mt_downloads_lc <- bind_rows(mt_downloads_lc)


```

### Convert to daily

```{r}


  nwm_daily <- mt_downloads_lc %>%
    mutate(predict_date = as_date(predict_dateTime)) %>%
    dplyr::group_by(comid, init_date, member, predict_date) %>%
    summarise(forecasted_q_cms = mean(modeled_q_cms)) %>%
    dplyr::ungroup()


```

### Take mean of all members and convert flow to cms/km2 

######## Normalize by drainage area

```{r}
##### Combine all members and take the mean
##### And transform discharge by normalizing by drainage area
##### And taking the log

nwm_daily_mean_forecast_lc_tribs <- nwm_daily %>%
  mutate(comid = as.character(comid)) %>%
  dplyr::group_by(comid, init_date, predict_date) %>%
  summarise(mean_forecasted_q_cms = mean(forecasted_q_cms)) %>%
  dplyr::ungroup() %>%
  dplyr::group_by(comid) %>%
  dplyr::arrange(init_date, .by_group = TRUE) %>%
  dplyr::ungroup() %>%
  mutate(lead_days = as.numeric(predict_date - init_date)) %>%
  inner_join(., lc_gages_metadata_clean %>%
               rename(drnarea_km2 = drain_area_km2) %>%
               dplyr::select(tributary, site_no, comid,drnarea_km2) %>%
               mutate(comid = as.character(comid)),
             by = "comid") %>%
  mutate(mean_forecasted_q_cms_km2 = mean_forecasted_q_cms/drnarea_km2) %>%
  mutate(mean_forecasted_log_q_cms_km2 = log10(mean_forecasted_q_cms_km2))


################################################################################
```


# Calculate various flow variables

### Antecedent conditions

This is mostly challenging because at longer leadtimes, "antecedent" conditions 
become a mixture of forecasted and observed values. So for, say, a three day lead time
weekly antecedent discharge day 1 would have four days of observed discharge (day 4,5,6,7)
and three days of forecasted discharge (1,2,3); on day 2, three days of observed and four days 
of forecasted, and so on and so forth. So we need to build a dataframe that reflects this

```{r, eval=FALSE}

### First, make sure the observed data is arranged by date and station

observed_flow_for_nwm_predictions <- flow_data_july2024_transform %>%
  dplyr::group_by(tributary) %>%
  arrange(date, .by_group = TRUE) %>%
  dplyr::ungroup()

#### Create a lookup table to determine which timesteps 
#### For which we need antecedent values 

timesteps_needed <- nwm_daily_mean_forecast_lc_tribs %>%
      ungroup() %>%
      mutate(days_needed_from_gage_weekly_ant = 7 - lead_days) %>%
      mutate(days_needed_from_gage_weekly_ant = 
               ifelse(days_needed_from_gage_weekly_ant < 0, 0, 
                    days_needed_from_gage_weekly_ant)) %>%
      mutate(days_needed_from_nwm_weekly_ant = 7 - days_needed_from_gage_weekly_ant) %>%
      mutate(days_needed_from_gage_monthly_ant = 30 - lead_days) %>%
      mutate(days_needed_from_gage_monthly_ant = 
               ifelse(days_needed_from_gage_monthly_ant < 0, 0, 
                    days_needed_from_gage_monthly_ant)) %>%
      mutate(days_needed_from_nwm_monthly_ant = (30) - 
               days_needed_from_gage_monthly_ant) %>%
          dplyr::select( comid, lead_days, init_date, predict_date, 
                      days_needed_from_gage_weekly_ant, 
                      days_needed_from_nwm_weekly_ant,
                      days_needed_from_gage_monthly_ant, 
                      days_needed_from_nwm_monthly_ant,
                     ) %>%
  inner_join(., lc_gages_metadata_clean %>%
               mutate(comid = as.character(comid)) %>%
               dplyr::select(site_no, comid),
             by = "comid")
  

#### Find antecedent conditions for all LC gages for each forecasted data
#### in water year 2022-2023

flow_ants_lc <- purrr::pmap_dfr(list(predict_date = timesteps_needed$predict_date,
                          init_date = timesteps_needed$init_date,
                          site_no = timesteps_needed$site_no,
                          days_from_gage =
                            timesteps_needed$days_needed_from_gage_weekly_ant,
                          days_from_model =
                            timesteps_needed$days_needed_from_nwm_weekly_ant,
                monthly_days_from_gage =
                  timesteps_needed$days_needed_from_gage_monthly_ant,
                monthly_days_from_model =
                  timesteps_needed$days_needed_from_nwm_monthly_ant
                          ),
                .f = possibly(antecedent_calculator,
                      otherwise = tibble(mean_prior_weekly_log_q_cms_km2 = NA, 
                                         mean_prior_monthly_log_q_cms_km2 = NA)),
                
                obs_df = observed_flow_for_nwm_predictions,
                model_df = nwm_daily_mean_forecast_lc_tribs,
                .progress = TRUE)

######


  

```

### Delta Q

```{r, eval=FALSE}

#### Calculate the delta q

delta_q <- limb_getter(nwm_daily_mean_forecast_lc_tribs,
                         observed_flow_for_nwm_predictions)


```

### Combine antecedent and delta Q

```{r, eval=FALSE}

antecedent_flow <- bind_cols(nwm_daily_mean_forecast_lc_tribs, 
                             flow_ants_lc) %>%
  inner_join(., 
             delta_q,
             by = c("predict_date", "init_date", "tributary")) 

```

### Put all the flow conditions together
```{r, eval=FALSE}

nwm_flow_and_antecedents <- antecedent_flow %>%
  dplyr::select(!delta_daily_q_cat) %>%
  mutate(day_of_year = yday(predict_date)) %>%
  dplyr::mutate(season = case_when(day_of_year %in% seq(60,151,1) ~ "Spring",
                                   day_of_year %in% seq(152,243,1) ~ "Summer",
                                   day_of_year %in% seq(244,334,1) ~ "Fall",
                                   day_of_year > 334 | day_of_year < 60 ~ "Winter")) %>%
  mutate(water_year = add_waterYear(predict_date)) %>%
  dplyr::select(!day_of_year) %>%
  rename(log_daily_q = mean_forecasted_log_q_cms_km2,
         mean_prior_weekly_q = mean_prior_weekly_log_q_cms_km2,
         mean_prior_monthly_q = mean_prior_monthly_log_q_cms_km2,
         ) %>%
  dplyr::select(tributary, 
                init_date, predict_date, lead_days,
                log_daily_q, 
                mean_prior_weekly_q, mean_prior_monthly_q, lag_daily_q,
                delta_daily_q,
                season,
                water_year
                )

################################################################################

#### Read from file if already done

#nwm_flow_and_antecedents <- read_csv(here("input-data/nwm_flow_and_antecedents.csv"))


#################################################################################

```

####################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################################

# Functions

This function essentially calculates various antecedent conditions to be used in
a forecasting scenario. Importantly, this requires combining observations AND 
forecasted values for a given forecast timestep, meaning that, for example, forecasted flow 
for Oct 31st 2022 has a weekly antecedent discharge of seven observed values for a forecast 
with lead time of zero days (so the one issued Oct 31st at midnight); of six observed and one
forecasted for a forecast with a lead time of one day (so issued at Oct 30th); and so on. This is 
kinda complicated to create a dataframe that follows this format

```{r antecedent_calculator}

antecedent_calculator <- function(predict_date,
                                  init_date,
                                  site_no,
                              days_from_gage = NULL, 
                              days_from_model = NULL,
                              monthly_days_from_gage = NULL,
                              monthly_days_from_model = NULL,
                              obs_df = NULL,
                              model_df = NULL
                              ) {

  
########################## SET UP ##############################################
  
  ### Find the index of the dateTime of interest in the observed data (minus 1)
  
  date_index_obs <- which(obs_df$date == predict_date &
                                obs_df$site_no == site_no) 
  
  ### Find the index of thepredict  date
  
  date_index_model <- which(model_df$predict_date == predict_date & 
                              model_df$site_no == site_no & 
                              model_df$init_date == init_date
                              )
  
  ########## WEEKLY ANTECEDENTS ###################################################
  
  ### Calculate weekly antecedents based on values calculated in the 
  ### timesteps_needed step
  
    if (days_from_gage != 0){
          log_obs_flow <- obs_df %>%
            dplyr::ungroup() %>%
            dplyr::slice(((date_index_obs - days_from_model) - days_from_gage):
                           (date_index_obs - days_from_model - 1)) %>%
            .$log_daily_q
          
        
    } else{
        
      log_obs_flow <- NA
      
      } ### End log observed flow ifelse 

    if(days_from_model != 0 ){
      
          log_modeled_flow <- model_df %>%
            dplyr::ungroup() %>%
            dplyr::slice((date_index_model - days_from_model):(date_index_model -
                                                                  1)) %>%
            .$mean_forecasted_log_q_cms_km2
        
      
        }  else {
        
        log_modeled_flow <- NA
      }


    obs_and_modeled_flow <-  c(log_obs_flow, log_modeled_flow) %>%
      na.omit() 
      
      mean_prior_weekly_flow <- mean(obs_and_modeled_flow) %>%
            as_tibble() %>%
            rename(mean_prior_weekly_log_q_cms_km2 = 1)
  

  ################################################################################
    
  ############# MONTHLY ANTECEDENTS ##############################################

    if (monthly_days_from_gage > 0 & monthly_days_from_model <= 7){

          log_obs_flow_minus3 <- obs_df %>%
            dplyr::slice((date_index_obs - 30):(date_index_obs -
                                                       7)) %>%
            .$log_daily_q

          obs_and_modeled_flow_monthly <- c(obs_and_modeled_flow,
                                            log_obs_flow_minus3) %>%
            na.omit()

    } else if(monthly_days_from_model > 7) {

      log_modeled_flow_monthly <- model_df %>%
            dplyr::slice((date_index_model -
                            monthly_days_from_model):(date_index_model - 1)) %>%
            .$mean_forecasted_log_q_cms_km2

      log_observed_flow_monthly <- obs_df %>%
            dplyr::slice(((date_index_obs - monthly_days_from_model)
                          - monthly_days_from_gage):(date_index_obs -
                                                        monthly_days_from_model - 1)) %>%
            .$log_daily_q

      obs_and_modeled_flow_monthly <- c(log_modeled_flow_monthly,
                                        log_observed_flow_monthly) %>%
        na.omit()

    } else {

      obs_and_modeled_flow_monthly <- NA

    }
      
            mean_prior_monthly_flow <- mean(obs_and_modeled_flow_monthly) %>%
            as_tibble() %>%
            rename(mean_prior_monthly_log_q_cms_km2 = 1)
            
            weekly_and_monthly_ant <- bind_cols(mean_prior_weekly_flow, 
                                                mean_prior_monthly_flow)


}
```



This function calculates the change in discharge from time t-1 to time t. It 
calculates this as both a raw numeric and then a categorical variable with
three possible values of -1, 0, and 1.


```{r limb_getter}

limb_getter <- function(model_df, observed_df){
  
  
  model_df %>%
    mutate(init_date_minus_one = init_date - days(1)) %>%
    group_by(init_date, tributary, site_no, comid) %>% #### This is to calculate limb
    mutate(delta_q_modeled = mean_forecasted_q_cms_km2  - 
             lag(mean_forecasted_q_cms_km2 )) %>%
    mutate(modeled_lag_daily_q = lag(mean_forecasted_log_q_cms_km2
                                     )) %>%
    mutate(delta_q_q_modeled = 
             abs(delta_q_modeled/lag(mean_forecasted_q_cms_km2))*sign(delta_q_modeled)) %>%
    inner_join(., observed_flow_for_nwm_predictions %>%
                 dplyr::select(tributary, site_no,
                               date, log_daily_q),
               join_by(init_date_minus_one == date,
                       tributary == tributary,
                       site_no == site_no)) %>%
    mutate(delta_q_day_zero = mean_forecasted_q_cms_km2 - 10^log_daily_q) %>%
    mutate(delta_q_q_day_zero = abs(delta_q_day_zero/10^log_daily_q)*sign(delta_q_day_zero)) %>%
    mutate(delta_q_q = coalesce(delta_q_q_modeled, delta_q_q_day_zero)) %>%
    group_by(init_date, tributary, site_no, comid) %>% #### This is to calculate limb
    mutate(delta_q_modeled = mean_forecasted_log_q_cms_km2 - 
             lag(mean_forecasted_log_q_cms_km2)) %>% ## Update
    mutate(delta_q_day_zero = mean_forecasted_log_q_cms_km2 - 
             log_daily_q) %>% ### Replace w/ log
    mutate(delta_q = coalesce(delta_q_modeled, delta_q_day_zero)) %>%
    rename(delta_daily_q = delta_q) %>%
    mutate(lag_daily_q = coalesce(modeled_lag_daily_q, log_daily_q)) %>%
    dplyr::ungroup() %>%
    mutate(delta_daily_q_cat = case_when(delta_q_q > 0.10 ~ 1,
                                       delta_q_q < -0.10 ~ -1,
                                       (delta_q_q <= 0.10 & delta_q_q >= -0.10) ~ 0)) %>%
    mutate(delta_daily_q_cat = as.factor(delta_daily_q_cat)) %>%
    dplyr::select(c(predict_date, init_date, tributary,
                    delta_daily_q, 
                    delta_daily_q_cat,
                    lag_daily_q))
    
  
}
```

