---
title: "02_toy_forecasting_example"
author: "JTK"
date: "2025-05-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

################################################################################

This script takes the models we built and trained in the first part (see [**01_model_development**](https://github.com/jtkemper/wqforecast-tutorial/blob/main/notebooks/01_model_development.ipynb)) to forecast water quality (specifically, total phosphorus concentration) in the Lake Champlain basin. 

To do so, it takes NWM operational forecast outputs (medium-term forecasts) for June and July 2024 to provide a toy example of forecasting water quality using the NWM.

We can use the three model "types" built in the last exercise to get a sort of ensemble forecast. 


**Inputs**

1) `nwm_flow_and_antecedents` (*from .csv*): operational NWM forecasts for June & July 2024 and a few engineered features from those forecasts (such as antecedent conditions)

2) `lgbm_`: trained LightGBM model architectures for three model types

  - `simplest`: model trained using observed concentration and observed discharge
  - `mrcomp`  model trained using observed concentration , discharge, and antecedent flow conditions 
  - `mstcomp` model trained using observed concentration, discharge,  antecedent flow conditions, and a number of static watershed features  

3) `watershed_chars`: a dataframe of static watershed attributes for each tributary

4) `trib_tp_conc` & `obs_flow_july2024`: dataframes containing observed discharge and water quality data for June & July 2024

And if [**01_model_development**](https://github.com/jtkemper/wqforecast-tutorial/blob/main/notebooks/01_model_development.ipynb) has not been run:

5) `mrcomp_chosen_mod` & `mstcomp_chosen_mod`: dataframe containing the model features from the feature selection process

**Outputs**

1) Operational-scale total phosphorus concentration forecasts for 18 tributaries of Lake Champlain for June & July 2024


################################################################################


# Housekeeping

### Install required packages

```{r}

### Uncomment if you have not yet run 01_model_development

#install.packages("here")

```

### Load packages

```{r}

require(here)

require(lightgbm)

require(tidyverse)

require(ggthemes)
```


# Get observed water quality & flow data

```{r}

### TP data

trib_tp_conc <- read_csv(here("input-data/toy-forecasting/trib_tp_conc_JunJuly2024.csv")) %>%
  filter(date > "2024-06-20")

### And flow data

obs_flow_july2024 <- read_csv(here("input-data/toy-forecasting/flow_data_july2024.csv"))

```


# Load other essential files

```{r}
### Get watershed characteristics 

watershed_chars <- read_csv(here("input-data/toy-forecasting/watershed_characteristics.csv"))

################################################################################################

### Get model features (if you ****haven't**** completed model selection and training)

mrcomp_chosen_mod <- read_csv(here("input-data/toy-forecasting/mrcomp_chosen_mod.csv"))

mstcomp_chosen_mod <- read_csv(here("input-data/toy-forecasting/mstcomp_chosen_mod.csv"))

##################################################################################################

################################################################################################

### Get model features (if you ***have*** completed model selection and training)

#mrcomp_chosen_mod <- read_csv(here("output-data/mrcomp_chosen_mod.csv"))

#mstcomp_chosen_mod <- read_csv(here("output-data/mstcomp_chosen_mod.csv"))

################################################################################################

### Get forecasting flow data

nwm_flow_and_antecedents <- read_csv(here("input-data/toy-forecasting/nwm_flow_and_antecedents.csv"))

### Get gage metadata

lc_gages_metadata_clean <- read_csv(here("input-data/toy-forecasting/lc_gages_metadata_clean.csv"))
```

# Create dataframes for forecasting

```{r}
### Create dataframes for prediction 

simplest_forecasting_df <- nwm_flow_and_antecedents %>%
  dplyr::select(tributary, init_date, predict_date, lead_days, log_daily_q) %>%
  dplyr::select(order(colnames(.)))

mrcomp_forecasting_df <- nwm_flow_and_antecedents %>%
  inner_join(., watershed_chars,
             by = "tributary") %>%
  dplyr::select(tributary, init_date, predict_date, lead_days, 
                mrcomp_chosen_mod$Feature) %>%
  dplyr::select(order(colnames(.)))

mstcomp_forecasting_df <- nwm_flow_and_antecedents %>%
  inner_join(., watershed_chars,
             by = "tributary") %>%
  dplyr::select(tributary, init_date, predict_date, lead_days, 
                mstcomp_chosen_mod$Feature) %>%
  dplyr::select(order(colnames(.)))

```

# Predict

### Load in models

```{r}
### Load in models

#### Find directory 

all_models <- fs::dir_ls(here("input-data/models"))

#### Load them in

lgbm_simplest <- all_models[str_detect(all_models, "simplest")]

lgbm_mrcomp <- all_models[str_detect(all_models, "mrcomp")]

lgbm_mstcomp <- all_models[str_detect(all_models, "mstcomp")]

```

### Forecast with each model

```{r}
### Forecast

#### Simplest model

simplest_forecasts <- lgbm_predictor(predictors_df = simplest_forecasting_df,
                                     observed_df = trib_tp_conc,
                                     trained_model = lgbm_simplest)

simplest_forecasts_ts <- simplest_forecasts[[3]] %>% mutate(model_typ = "simplest")

#### More complex model

mrcomp_forecasts <- lgbm_predictor(predictors_df = mrcomp_forecasting_df,
                                     observed_df = trib_tp_conc,
                                     trained_model = lgbm_mrcomp)

mrcomp_forecasts_ts <- mrcomp_forecasts[[3]] %>% mutate(model_typ = "mrcomp")

#### Most complex model

mstcomp_forecasts <- lgbm_predictor(predictors_df = mstcomp_forecasting_df,
                                     observed_df = trib_tp_conc,
                                     trained_model = lgbm_mstcomp)

mstcomp_forecasts_ts <- mstcomp_forecasts[[3]] %>% mutate(model_typ = "mstcomp")

```

# Plot forecasts

```{r}

### Log axis breaks ticks
### Log breaks

log_breaks <- 10^(-10:10)

log_minor_breaks <- rep(1:9, 21)*(10^rep(-10:10, each=9))

### Forecasted concentration

all_forecast_ts <- bind_rows(simplest_forecasts_ts,
                             mrcomp_forecasts_ts,
                             mstcomp_forecasts_ts)

winoo_tp <- trib_tp_conc %>%
                 filter(tributary == "Winooski River")

all_forecast_ts %>%
  dplyr::filter(tributary == "Winooski River") %>%
  ggplot() +
    geom_line(aes(x = predict_date, y = forecasted_conc, 
                  color = model_typ), linewidth = 1) +
    geom_point(data = trib_tp_conc %>%
                 filter(tributary == "Winooski River"), 
               aes(x = date, y = 10^log_conc),
               shape = 2, size = 1, color ="black") +
    scale_color_brewer(palette = "Set2") +
    scale_x_date(date_breaks = "month",
                     date_minor_breaks = "week") + 
    scale_y_log10(breaks = log_breaks,
                  minor_breaks = log_minor_breaks) + 
    labs(x = element_blank(),
         y = "TP Conc. (µg/L)") +
    theme_few() +
    theme(panel.grid = element_line(color = "gray90"),
          legend.position = "bottom") +
    facet_wrap(~init_date, nrow = 4)

#### Forecasted flow

forecasted_and_obs_flow <- inner_join(nwm_flow_and_antecedents %>%
                                        dplyr::select(!c(water_year, season)) %>%
                                        mutate(daily_q = 10^log_daily_q) %>%
                                        rename_with(~paste0(., "_forc"), where(is.numeric)),
                                      obs_flow_july2024 %>%
                                        mutate(daily_q = 10^log_daily_q),
                                      join_by(predict_date == date,
                                              tributary == tributary))


forecasted_and_obs_flow %>%
  dplyr::select(tributary, init_date, predict_date, daily_q, daily_q_forc) %>%
  pivot_longer(cols = c(daily_q, daily_q_forc), 
               values_to = "flow_cms_km2",
               names_to = "typ") %>%
  mutate(src = ifelse(str_detect(typ, "forc"), "NWM", "Observed - USGS")) %>%
  inner_join(., lc_gages_metadata_clean %>%
               dplyr::select(tributary, drain_area_km2)) %>%
  mutate(flow_cms = flow_cms_km2*drain_area_km2) %>%
  dplyr::filter(tributary == "Winooski River") %>%
  ggplot() +
    geom_line(aes(x = predict_date, y = flow_cms, 
                  color = src), linewidth = 1) +
    scale_color_manual(values = c("darkblue", "dodgerblue")) + 
    scale_x_date(date_breaks = "month",
                     date_minor_breaks = "week") + 
    scale_y_log10(breaks = log_breaks,
                  minor_breaks = log_minor_breaks) + 
    labs(x = element_blank(),
         y = "TP Conc. (µg/L)") +
    theme_few() +
    theme(panel.grid = element_line(color = "gray90"),
          legend.position = "bottom") +
    facet_wrap(~init_date, nrow = 4)





```


################################################################################
################################################################################
################################################################################
################################################################################

# Functions


```{r lgbm_predictor}

lgbm_predictor <- function(predictors_df = NULL ,
                           observed_df = NULL,
                           trained_model = NULL){
  
  ### Empty list to store stuff
  
  nwm_error_and_full_ts <- list()
  
  ### Set up the data to use with lightgbm
  

     forecast_data <- data.matrix(predictors_df %>%
                                  dplyr::select(!c(init_date, 
                                                   lead_days, 
                                                   predict_date,
                                                   tributary)) %>%  
                               dplyr::select(order(colnames(.))))


  #### Load in the tuned LGBM model
  
        message(crayon::green(paste("Loading model ", trained_model, "\n")))

    
        final_tuned_model <- lgb.load(trained_model)
      

  
  ### Forecast with the model, bind to other relevant data
  
        forecast_from_nwm <- predict(final_tuned_model,
                                     newdata = forecast_data) %>%
          as_tibble() %>%
          rename(log_forecasted_conc = 1) %>%
          bind_cols(., predictors_df %>%
                      rename_with(~paste0(., "_forc"), where(is.numeric))) %>%
          rename(log_modeled_flow =
                      log_daily_q_forc,
                 lead_days = lead_days_forc)

    ### Now bind to observations
        
          forecast_and_obs <- forecast_from_nwm %>%
            inner_join(., observed_df %>%
                   dplyr::select(tributary,
                                 date,
                                 log_conc),
                   join_by(predict_date == date,
                           tributary == tributary)) %>%
          rename(log_observed_conc = log_conc) 


        ### Transform back to linear
            forecast_and_obs <- forecast_and_obs %>%
                    mutate(forecasted_conc = 10^log_forecasted_conc) %>%
                    mutate(observed_conc = 10^log_observed_conc)

            nwm_error_and_full_ts[[1]] <- forecast_and_obs 

            ### Calculate some errors
            
           nwm_forecast_error <-  forecast_and_obs %>%
                    dplyr::group_by(tributary, lead_days) %>%
                    summarise(
                      mae = hydroGOF::mae(forecasted_conc, observed_conc)) 



            nwm_error_and_full_ts[[2]] <- nwm_forecast_error 
            
            forecasted_contin_ts <- forecast_from_nwm %>%
              mutate(forecasted_conc = 10^log_forecasted_conc) %>%
              dplyr::select(tributary,
                            init_date, predict_date, lead_days,
                            forecasted_conc)
            
            
       

            nwm_error_and_full_ts[[3]] <- forecasted_contin_ts 


            return(nwm_error_and_full_ts)
            
            
}


```




